---
title: "R Notebook"
output: html_notebook
---
# loading library and data
```{r}
library(data.table)
library(tidyverse)

df <- fread("./ProcessedData/Train.csv")
```

# define gun violence per 10,000 people
```{r}
df <- df %>%
  mutate(`Population 2020` = readr::parse_number(`Population 2020`)) %>%
  mutate(vioIndex = n_totVictim / `Population 2020` * 10000) %>%
  select(!c(Code, `Population 2020`))

dfModelReady <- df %>%
  select(!c(state, city_or_county, n_killed, n_injured, n_totVictim,
            povRate_19, povRate_minor_19, median_income_19))
```

# divide df into differnet years
```{r}
#2018
df18 <- dfModelReady %>%
  filter(IncYear == 2018) %>%
  select(!c(IncYear, UnempRate2020, UnempRate2019, UnempRate2018))

#2019
df19 <- dfModelReady %>%
  filter(IncYear == 2019) %>%
  select(!c(IncYear, UnempRate2020, UnempRate2019, UnempRate2017))

#2020
df20 <- dfModelReady %>%
  filter(IncYear == 2020) %>%
  select(!c(IncYear, UnempRate2020, UnempRate2018, UnempRate2017))

#2021
df21 <- dfModelReady %>%
  filter(IncYear == 2021) %>%
  select(!c(IncYear, UnempRate2019, UnempRate2018, UnempRate2017))

#2022
df22 <- dfModelReady %>%
  filter(IncYear == 2022) %>%
  select(!c(IncYear, UnempRate2019, UnempRate2018, UnempRate2017))
```

# descriptive analysis for dependent / Outcome varaible (gun violence)
```{r}
# check distribution of victim
df %>%
  ggplot(aes(x = n_totVictim)) +
  geom_density() +
  ggtitle('Distribution of Victim Count (County level)') +
  xlab('Total Victim (County Level)') 

# histogram
df %>%
  ggplot(aes(x = n_totVictim)) +
  geom_bar() +
  ggtitle('Histogram of Victim Count') +
  xlab('Total Victim (County Level)') 

# by year
df %>%
  ggplot(aes(x = n_totVictim)) +
  geom_bar() +
  facet_grid(IncYear~.) +
  ggtitle('Victim Count Distribution by Year') +
  xlab('Total Victim (County Level)') 
```

**Comment**: Highly skewed data

```{r}
# check distribution of victim per 10,000 people
df %>%
  ggplot(aes(x = vioIndex)) +
  geom_density() +
  ggtitle('Distribution of Victim Count per 10,000 People') +
  xlab('Victims of Gun Violence per 10,000 People (County Level)')
  

# by year
df %>%
  ggplot(aes(x = vioIndex)) +
  geom_density() +
  facet_grid(IncYear~.)
```

**Comment**: victim per 10,000 people is also skewed, looks looks like lots of county already have gun violence incidents this year.

# descriptive analysis for main explantory varaible (poverty)
```{r}
# check distribution of poverty rate 2019
df %>%
  ggplot(aes(x = povRate_19)) +
  geom_density() +
  ggtitle('Distribution of 2019 Poverty Rate') +
  xlab('2019 Poverty Rate (County Level)')

# histogram
df %>%
  ggplot(aes(x = povRate_19)) +
  geom_histogram(bins = 40) +
  ggtitle('Histogram of 2019 Poverty Rate') +
  xlab('2019 Poverty Rate (County Level)')
```

**Comment**: Not normal

```{r}
# check distribution of poverty rate 2015 to 2019
df %>%
  ggplot(aes(x = poveRate_15_19)) +
  geom_density() +
  ggtitle('Distribution of Average Poverty Rate from 2015 to 2019') +
  xlab('2015 - 2019 Average Poverty Rate (County Level)')

df %>%
  ggplot(aes(x = poveRate_15_19)) +
  geom_histogram(bins = 40) +
  ggtitle('Histogram of Average Poverty Rate from 2015 to 2019') +
  xlab('2015 - 2019 Average Poverty Rate (County Level)')
```

**Comment**: kind normal

# fit a model for 2022
```{r}
model22 <- lm(log(vioIndex) ~ . , data = df22)
summary(model22)
```

## try feature selection to find best control varaibles
```{r}
# set an empty model
none <- lm(log(vioIndex) ~ 1, data = df22)

# find MSE
MSE <- (summary(model22)$sigma)^2
step(none, scope = list(upper = model22),scale = MSE,direction="both")
```

## refit mode with selected control variables 2022
```{r}
model22_select <- lm(formula = log(vioIndex) ~ poveRate_15_19 + deepPov_minor + highSch_15_19 + UnempRate2020 + LFPVetsRate, data = df22)
summary(model22_select)

plot(model22_select,1)
plot(model22_select,2)
```


# fit a model for 2021
```{r}
model21 <- lm(log(vioIndex) ~ . , data = df21)
summary(model21)
```

## find best control variables
```{r}
# set an empty model
none <- lm(log(vioIndex) ~ 1, data = df21)

# find MSE
MSE <- (summary(model21)$sigma)^2
step(none, scope = list(upper = model21),scale = MSE,direction="both")
```

## refit model 2021
```{r}
model21_select <- lm(formula = log(vioIndex) ~ poveRate_15_19 + deepPov_minor + highSch_15_19 + UnempRate2020 + LFPVetsRate, data = df21)
summary(model21_select)

plot(model21_select,1)
plot(model21_select,2)
```

**comment**: UnempRate2020 + highSch_15_19 + LFPVetsRate always explain violence and looks like it is poverty in minor rather than pverty of all explains violent

# fit a model for 2020
```{r}
model20 <- lm(log(vioIndex) ~ . , data = df20)
summary(model20)
```

## find best control variables
```{r}
# set an empty model
none <- lm(log(vioIndex) ~ 1, data = df20)

# find MSE
MSE <- (summary(model20)$sigma)^2
step(none, scope = list(upper = model20),scale = MSE,direction="both")
```

## refit model 2020
```{r}
model20_select <- lm(formula = log(vioIndex) ~ poveRate_15_19 + deepPov_minor + highSch_15_19 + UnempRate2019 + 
    LFPVetsRate,
    data = df20)
summary(model20_select)

plot(model20_select,1)
plot(model20_select,2)

library(Stat2Data)
cooksplot(model20_select)
```

**comment**: again, highSch_15_19 + UnempRate2019 + LFPVetsRate is very important contorl variables, poverty explains stuff, but poverty in minor is more important.

## delete outliers
```{r}
#delete outliers
df20 <- df20[-c(5,13,19)]
model20_select <- lm(formula = log(vioIndex) ~ poveRate_15_19 + deepPov_minor + highSch_15_19 + UnempRate2019 + LFPVetsRate, data = df20)
summary(model20_select)

plot(model20_select,1)
plot(model20_select,2)

cooksplot(model20_select)
```


## make 2020's model looks better (more organized)
```{r}
library(stargazer)
stargazer(model20_select, type = "text",
          title = "Linear Model for 2020 Gun Violence (County Level)",
          dep.var.labels = "Log Number of Victims per 10,000 people",
          covariate.labels = c("Average poverty rate of last 5 year (2015 - 2019)",
                               "Average deep poverty rate for children of last 5 years (2015 - 2019)",
                               "Average proportion of population has high school diploma of last 5 years (2015 - 2019)",
                               "Unemployment rate of last year (2019)",
                               "Average labor force participation rate for veterans age 18 to 64 in last 5 years (2015 - 2019)",
                               "Intercept"))
```

## plot for 2020
```{r}
library(sjPlot)

plot_model(model20_select, type = "pred",
           terms = "poveRate_15_19",
           show.data = TRUE,
           title = 'Predicted Victims per 10,000 people',
           axis.title = c('Average Poverty Rate of Last 5 Year (2015 - 2019)',
                          'Victims per 10,000 People (Back Transformed)'))

plot_model(model20_select, type = "pred",
           terms = "deepPov_minor",
           show.data = TRUE,
           title = 'Predicted Victims per 10,000 people',
           axis.title = c('Average Deep Poverty Rate for Children of Last 5 Years (2015 - 2019)',
                          'Victims per 10,000 People (Back Transformed)'))
```


# fit a model for 2019
```{r}
model19 <- lm(log(vioIndex) ~ . , data = df19)
summary(model19)
```


## find best control variables
```{r}
# set an empty model
none <- lm(log(vioIndex) ~ 1, data = df19)

# find MSE
MSE <- (summary(model19)$sigma)^2
step(none, scope = list(upper = model19),scale = MSE,direction="both")
```

## refit model for 2019
```{r}
model19_select <- lm(formula = log(vioIndex) ~ poveRate_15_19 + deepPov_minor + highSch_15_19 + UnempRate2018 + 
    LFPVetsRate,
    data = df19)
summary(model19_select)
```

**comment**: highSch_15_19 + PopChangeRate1019 explains

# fit a model for 2018
```{r}
model18 <- lm(log(vioIndex) ~ . , data = df18)
summary(model18)
```

## refit model 2018
```{r}
model18_select <- lm(formula = log(vioIndex) ~ poveRate_15_19 + deepPov_minor + highSch_15_19 + UnempRate2017 + 
    LFPVetsRate,
    data = df18)
summary(model18_select)
```


#conclusion
poverty not necessary show strong evidence in every year, but poverty in minority show strong evidence of influence. Moreover, education (percent own high school diploma), unemployment rate of last year, and Labor force participation rate for veterans age 18 to 64, 2015-19 also show very consistent influence over the gun violence.
